Using high-resolution variant frequencies to empower clinical genome interpretation

### 2016-12-16 manuscript/SupplementaryMaterial.Rmd
Responding to reviewers comments

@@

The prevalence of a recessive condition can be related to the AF of causative variants by:  
*Prevalence =* $\sum$ *(AF of causative alleles in each contributing gene)^2^ x penetrance*  
approximating to:   
*Prevalence = (combined frequency of causative alleles in gene)^2^ x (number of similar genes) x penetrance*  
and expanding to:  
*Prevalence = (max individual AF x 1/maximum allelic contribution)^2^ x 1/maximum genetic contribution x penetrance*

where *maximum genetic contribution* represents the proportion of all cases that are attributable to the gene under evaluation, and *maximum allelic contribution* represents the proportion of cases attributable to that gene that are attributable to an individual variant. This can be rearranged to give the maximum frequency of a recessive disease causing variant in the population as:
*max credible AF* = $\sqrt($*prevalence*$) \times$ *maximum allelic contribution* $\times \sqrt($*maximum genetic contribution*$) \times 1/\sqrt($*penetrance*$)$

@@

It is often difficult to obtain accurate penetrance information for reported variants, and it is also difficult to know what degree of penetrance to expect or assume in newly discovered pathogenic variants. In this work we uniformly apply a value of 50% penetrance for inherited cardiac conditions (i.e. assuming variant penetrance is no lower than 50%) equivalent to that reported for our HCM example variant, and the lowest we found reported for any of our examples or reported accross our studied disorders. We recognise several other approaches that can be used to deal with the issues of penetrance, these include: setting a penetrance level equivalent to the minimum that is 'clinically actionable' for a disorder; lowering the penetrance if reduced penetrance is expected in a family; or using a two tiered approach, initially searching for a high-moderate penetrance variant but allowing for a lower-penetrance variant in a second pass. We believe that the ease of re-calculating our "maximum credible population allele frequency" lends itself to any of these approaches. We provide an online calculator to facilitate the exploration of these parameters (<https://jamesware.shinyapps.io/alleleFrequencyApp/>). If there are large case and control populations for a disease and the diease prevalence is known, we can use these to estimate penetrance `r citep("10.1126/scitranslmed.aad5169")`.

@@

We also note that occasionally a variant is seen in individuals falling under the Finnish or "Other" population categories in ExAC, and is a singleton or absent in all five continental populations. For these variants, the filtering allele frequency is set to zero. Because the Finnish are a bottlenecked population, disease-causing alleles may reach frequencies that would be impossible in large outbred populations. Similarly, because we have not assigned ancestry for the "Other" individuals, it is difficult to assess the population frequency of variants seen only in this set of individuals. Users are left to judge whether variants that  would not be filtered on the basis of frequency in the five continental populations, but that are recurrent in Finnish or "Other" populations, should be removed from consideration according to the specific circumstances.

@@

@@

Variants previously reported as causative of HCM either in ClinVar, or in a clinical series of 6179 HCM cases, but that were observed in ExAC above the maximum tolerated allele count for HCM (AC>9 globally) were manually curated according to the ACMG guidelines for interpretation and classification on sequence variants. *These variants were previously categorised as Likely Pathogenic in the LMM and Oxford HCM cohorts.

>>>>

For dominant disorders we define the maximum credible population AF (for a pathogenic allele) as:

*maximum credible population AF = prevalence* $\times$ *maximum allelic contribution* $\times$ *1/penetrance*

where _maximum allelic contribution_ is the maximum proportion of cases potentially attributable to a single allele, a measure of heterogeneity. It is important to note that although we estimate the _maximum allelic contribution_ as a single value, this can be broken down into seperate factors corresponding to the maximum contribution of any one gene (_maximum genetic contribution_) and the _maximum allelic contribution_ within that gene. These two seperate factors are used in the derivation of the recessive model described below.

@@

For a biallelic condition caused by only a single gene, the prevalence can be determined from the combined frequency of all possible pairs of causative alleles in that gene:

$prevalence = \sum_{i}^{j} (AF_{i} \times AF_{j} \times penetrance_{i,j})$

where $AF_{i}$ and $AF_{j}$ represent the frequencies of the two alleles, and $i$ and $j$ may be the same allele.

For a condition caused by multiple genes, the above equation must then be summed across $n$ contributing genes:

$prevalence = \sum_{1}^{n} \sum_{i}^{j} (AF_{i} \times AF_{j} \times penetrance_{i,j})$

Since the penetrance can not be separately estimated for each combination of alleles (though is typically high for recessive conditions), we simplify by estimating penetrance as constant for the condition, reducing the equation to:

$prevalence = penetrance \times \sum_{1}^{n}\sum_{i}^{j} (AF_{i} \times AF_{j})$

which is equivalent to:

$prevalence = penetrance \times \sum_{1}^{n} \sum_{i}^{} (AF_{i}^2)$

we can rearrange this to give an upper bound for the maximum credible AF of an individual causative variant:

$max$ $credible$ $AF$ = $\sqrt(prevalence) \times$ $maximum$ $allelic$ $contribution$ $\times \sqrt(maximum$ $genetic$ $contribution$) $\times 1/\sqrt(penetrance)$

where rather than simply dividing by n genes to get an average, we take the upper bound of the proportion attributable to any single gene (the *maximum genetic contribution*). Similarly we take the maximum proportion of cases attributable to that gene that are attributable to any individual variant (the *maximum allelic contribution*).

@@

It is often difficult to obtain accurate penetrance information for reported variants, and it is also difficult to know what degree of penetrance to expect or assume in newly discovered pathogenic variants. In this work we uniformly apply a value of 50% penetrance for inherited cardiac conditions (i.e. assuming variant penetrance is no lower than 50%) equivalent to that reported for our HCM example variant, and the lowest we found reported for any of our examples or reported accross our studied disorders. We recognise several other approaches that can be used to deal with the issues of penetrance, these include: setting a penetrance level equivalent to the minimum that is 'clinically actionable' for a disorder; lowering the penetrance if reduced penetrance is expected in a family; or using a two tiered approach, initially searching for a high-moderate penetrance variant but allowing for a lower-penetrance variant in a second pass. We believe that the ease of re-calculating our "maximum credible population allele frequency" lends itself to any of these approaches. We provide an online calculator to facilitate the exploration of these parameters (<http://cardiodb.org/alleleFrequencyApp>). If there are large case and control populations for a disease and the diease prevalence is known, we can use these to estimate penetrance `r citep("10.1126/scitranslmed.aad5169")`.

@@

We also note that occasionally a variant is seen in individuals falling under the Finnish or "Other" population categories in ExAC, and is a singleton or absent in all five continental populations. For these variants, the filtering allele frequency is set to zero. Because the Finnish are a bottlenecked population, disease-causing alleles may reach frequencies that would be impossible in large outbred populations. Similarly, because we have not assigned ancestry for the "Other" individuals, it is difficult to assess the population frequency of variants seen only in this set of individuals. Users are left to judge whether variants that  would not be filtered on the basis of frequency in the five continental populations, but that are sufficiently frequent in Finnish or "Other" populations, should be removed from consideration according to the specific circumstances.

@@

### Analysis tools

R was used for all analyses, as well as to compile the figures, manuscript and supplementary material, making use of the following packages:
knitr/rmarkdown (https://github.com/yihui/knitr), knitcitations (http://www.carlboettiger.info/2012/05/30/knitcitations.html), knitauthors (https://github.com/jamesware/knitauthors), dplyr (https://github.com/hadley/dplyr), ggplot2 (http://ggplot2.org/), packrat (https://rstudio.github.io/packrat/).

@@

Variants previously reported as causative of HCM either in ClinVar, or in a clinical series of 6179 HCM cases, but that were observed in ExAC above the maximum tolerated allele count for HCM (AC>9 globally) were manually curated according to the ACMG guidelines for interpretation and classification on sequence variants. For 5/43 variants, ClinVar contained the only identifiable disease-associated variant record, while 38/43 had additional information in the literature. Curated variant interpretations are deposited in ClinVar under the submission name "HCM\_ExAC\_frequency\_review\_2016". *These variants were previously categorised as Likely Pathogenic in the LMM and Oxford HCM cohorts.

### 2016-12-16 manuscript/manuscript.Rmd
Responding to reviewers comments

```{r authors, results='asis'}
read.delim("../data-raw/authors.txt") %>%
  formatAuthors
```

@@

Disease prevalence estimates were obtained from the literature and taken as the highest value reported. Cardiovascular diseases variants were modelled with a pentrance of 0.5, corresponding to the reported penetrance of the HCM variant used to illustrate our approach `r citep("10.1161/CIRCRESAHA.109.216291")` and the minimum found accross a range of variants/disorders.

@@

Pathogenic and non-conflicted variants were extracted from ClinVar (July 9, 2015 release) as described previously `r citep("10.1038/nature19057")`. ExAC counts were determined by matching on chromosome, position, reference, and alternate alleles. For all variants above the proposed maximum tolerated allele count for HCM, HGMD literature was reviewed and the level of evidence supporting pathogenicity was curated according to ACMG criteria `r citep("10.1038/gim.2015.30")`.

@@

322 HCM patients (confirmed by cardiac MRI) recruited to the NIHR Royal Brompton cardiovascular BRU were sequenced using the IlluminaTruSight Cardio Sequencing Kit `r citep("10.1007/s12265-016-9673-5")` on Illumina MiSeq and NextSeq platforms. This study had ethical approval (REC: 09/H0504/104+5) and informed consent was obtained for all subjects. The number of rare variants in _MYBPC3_, _MYH7_ and the six other sarcomeric genes associated with HCM (_TNNT2_, _TNNI3_, _MYL2_, _MYL3_, _TPM1_ and _ACTC1_) were calculated for this HCM cohort, and for reference population samples from ExAC. Variant frequencies were calculated for all protein altering variants (frameshift, nonsense, splice donor/acceptor, missense and in-frame insertions/deletions), with case/control odds ratios calculated separately for non-overlapping ExAC AF bins with the following breakpoints: 1x10^-5^, 5x10^-5^, 1x10^-4^, 5x10^-4^ and 1x10^-3^. Odds Ratios were calculated as OR = (cases with variant/cases without variant) / (ExAC samples with variant/ExAC samples without variant). In the absence of sample-level genotype data for ExAC, the number of samples with a variant was approximated by the total number of variant alleles (assuming each rare variant was from a distinct sample).

@@

To facilitate these calculations, we have produced an online calculator (<https://jamesware.shinyapps.io/alleleFrequencyApp/>) that will compute maximum credible population AF and maximum sample allele count for a user-specified genetic architecture, and conversely allow users to dynamically explore what genetic architecture(s) might be most compatible with an observed variant having a causal role in disease.

#### *Assessing the acuracy of our approach*

For all diseases with cases series that permitted us to define the genetic arcitecture, the commonest variant in the cases series was well within the calculated maximum alelle count in ExAC (Table 1).

@@

The above analysis applied a single global allele count limit of 9 for HCM, however, as AFs differ between populations, filtering based on frequencies in individual populations may provide greater power `r citep("10.1038/nature19057")`. For example, a variant relatively common in any one population is unlikely pathogenic, even if rare in other populations, provided the disease prevalence and architecture is consistent across populations. We therefore compute a maximum tolerated AC for each distinct sub-population of our reference sample, and filter based on the highest AF observed in any major continental population (see **Methods** and **Suplementary Note 3**).

@@

Finally, if the allelic heterogeneity of a disorder is not well characterised, it is conservative to assume minimal heterogeneity, so that the contribution of each gene is modelled as attributable to one allele, and the maximum allelic contribution is substituted by the maximum genetic contribution (i.e the maximum proportion of the disease attributable to single gene). For classic Ehlers-Danlos syndrome, up to 40% of the disease is caused by variation in the _COL5A1_ gene `r citep("10.1097/GIM.0b013e3181eed412")`. Taking 0.4 as our maximum allelic contribution, and a population prevalence of 1/20,000 `r citep("10.1097/GIM.0b013e3181eed412")` we derive a maximum tolerated ExAC AC of 5 (Table 2).

@@

Our method is designed to be complemenary to and used along side other gene and variant level methods to filter and prioritise candiate variants (e.g. gene level constraint`r citep("10.1038/ng.3050")`, amino acid conservation and missense prediction algorithms) along with segregation and functional data.

We recognise several limitations of our approach. First, we are limited by our understanding of the prevalence and genetic architecture of the disease in question: this characterisation will vary for different diseases and in different populations, though we illustrate approaches for estimation and extrapolation of parameters. In particular, we must be wary of extrapolating to or from less-well characterised populations that could harbour population-specific founder mutations. While incomplete knowledge of the genetic architecture of a disease of interest will limit this or any approach to evaluate a specific variant that has been observed at low frequency in a reference population, our framework and accompanying web tool do at least transparently define the range of disease architectures that are compatible with the observed data. For example, we may not be able to say definitively that a variant is benign, but we might be confident that it could only be causal if the prevalence of the disease is higher than published estimates, or its penetrance is below a specified value.

@@

All data required to reproduce the analysis, figures and manuscript (compiled in R) are available at <https://github.com/ImperialCardioGenetics/frequencyFilter>. Curated variant interpretations are deposited in ClinVar *Details to be added*. ExAC annotations are available at <https://exac.broadinstitute.org>. Our AF calculator app is located at <https://jamesware.shinyapps.io/alleleFrequencyApp/>, with source code available at <http://github.com/jamesware/alleleFrequencyApp>.

@@

The clinical utility of stringent allele frequency thresholds. (a) The number of predicted protein-altering variants (definition in Methods) per exome as a function of the allele frequency filter applied. A one-tailed 95% confidence interval is used, meaning that variants were removed from consideration if their AC would fall within the top 5% of the Poisson probability distribution for the user's maximum credible AF (x axis). (b) The odds ratio for HCM disease-association against allele frequency. The prevalence of variants in HCM-associated genes (_MYH7_, _MYBPC3_ and other sarcomeric  (_TNNT2_, _TNNI3_, _MYL2_, _MYL3_, _TPM1_ and _ACTC1_, analysed collectively)) in 322 HCM cases and 60,706 ExAC controls were compared for a range of allele frequency bins, and an odds ratio computed (see **Methods**). Data for each bin is plotted at the upper allele frequency cutoff. Error bars represent 95% confidence intervals. The probability that a variant is pathogenic is much greater at very low allele frequencies.

>>>>

```{r authors, results='asis'}
read.delim("../data-raw/authors.txt") %>%
  formatAuthors(infoText2="These authors jointly supervised this work")
```

@@

Disease prevalence estimates were obtained from the literature and taken as the highest value reported. Cardiovascular disease variants were modeled with a penetrance of 0.5, corresponding to the reported penetrance of the HCM variant used to illustrate our approach `r citep("10.1161/CIRCRESAHA.109.216291")` and the minimum found across a range of variants/disorders.

@@

Pathogenic and non-conflicted variants were extracted from ClinVar (July 9, 2015 release) as described previously `r citep("10.1038/nature19057")`. ExAC counts were determined by matching on chromosome, position, reference, and alternate alleles. For all variants above the proposed maximum tolerated allele count for HCM, literature from both HGMD and PubMed was reviewed and the level of evidence supporting pathogenicity was curated according to ACMG criteria `r citep("10.1038/gim.2015.30")`.

@@

322 HCM patients and 852 healthy volunteers (both confirmed by cardiac MRI) recruited to the NIHR Royal Brompton cardiovascular BRU were sequenced using the IlluminaTruSight Cardio Sequencing Kit `r citep("10.1007/s12265-016-9673-5")` on Illumina MiSeq and NextSeq platforms. This study had ethical approval (REC: 09/H0504/104+5) and informed consent was obtained for all subjects. The number of rare variants in the eight sarcomeric genes associated with HCM (_MYBPC3_, _MYH7_, _TNNT2_, _TNNI3_, _MYL2_, _MYL3_, _TPM1_ and _ACTC1_) were calculated for all protein altering variants (frameshift, nonsense, splice donor/acceptor, missense and in-frame insertions/deletions), with case/control odds ratios calculated separately for non-overlapping ExAC AF bins with the following breakpoints: 4x10^-5^, 1x10^-4^, 5x10^-4^ and 1x10^-3^. Odds Ratios were calculated as OR=(cases with variant/cases without variant)/(controls with variant/controls without variant).

@@

To facilitate these calculations, we have produced an online calculator (<http://cardiodb.org/alleleFrequencyApp>) that will compute maximum credible population AF and maximum sample allele count for a user-specified genetic architecture, and conversely allow users to dynamically explore what genetic architecture(s) might be most compatible with an observed variant having a causal role in disease.

#### *Assessing the accuracy of our approach*

For all diseases with cases series that permitted us to define the genetic architecture, the commonest variant in the cases series was well within the calculated maximum allele count in ExAC (Table 1).

@@

The above analysis applied a single global allele count limit of 9 for HCM, however, as AFs differ between populations, filtering based on frequencies in individual populations may provide greater power `r citep("10.1038/nature19057")`. For example, a variant relatively common in any one population is unlikely pathogenic, even if rare in other populations, provided the disease prevalence and architecture is consistent across populations. We therefore compute a maximum tolerated AC for each distinct sub-population of our reference sample, and filter based on the highest AF observed in any major continental population (see **Methods**).The tightness of the Poisson distribution used to compute maximum tolerated allele count is a function of sample size, and thus our approach is more conservative when allele number is lower, thus avoiding inappropriately filtering variants due to chance observation of a few alleles in a smaller sub-population or at a poorly genotyped site (see **Supplementary Note 3**).

@@

Finally, if the allelic heterogeneity of a disorder is not well characterised, it is conservative to assume minimal heterogeneity, so that the contribution of each gene is modeled as attributable to one allele, and the maximum allelic contribution is substituted by the maximum genetic contribution (i.e the maximum proportion of the disease attributable to single gene). For classic Ehlers-Danlos syndrome, up to 40% of the disease is caused by variation in the _COL5A1_ gene `r citep("10.1097/GIM.0b013e3181eed412")`. Taking 0.4 as our maximum allelic contribution, and a population prevalence of 1/20,000 `r citep("10.1097/GIM.0b013e3181eed412")` we derive a maximum tolerated ExAC AC of 5 (Table 2).

@@

Our method is designed to be complementary to and used along side other gene and variant level methods to filter and prioritise candiate variants (e.g. gene level constraint`r citep("10.1038/ng.3050")`, amino acid conservation`r citep(c("10.1101/gr.3715005","10.1101/gr.3577405"))` and missense prediction algorithms`r citep(c("10.1038/nmeth0410-248","10.1093/nar/gks539"))`) along with segregation and functional data.

We recognise several limitations of our approach. First, we are limited by our understanding of the prevalence and genetic architecture of the disease in question: this characterisation will vary for different diseases and in different populations, though we illustrate approaches for estimation and extrapolation of parameters. In particular, we must be wary of extrapolating to or from less-well characterised populations that could harbour population-specific founder mutations. While incomplete knowledge of the genetic architecture of a disease of interest will limit this or any approach to evaluate a specific variant that has been observed at low frequency in a reference population, our framework and accompanying web tool do at least transparently define the range of disease architectures that are compatible with the observed data. For example, many neurological disorders have Mendelian forms as well as idiopathic forms with genetic risk factors of modest effect sizes, high allelic and genetic heterogeneity, and/or dramatic variability in the penetrance of different variants `r citep(c("10.1016/j.cell.2013.10.037","10.1016/j.cell.2015.01.019","10.1016/j.cell.2015.01.052","10.1371/journal.pgen.1002548","10.1126/scitranslmed.aad5169"))`. Reference population allele frequency information alone can never definitively show that a variant possesses no association to disease, but it can still provide sensible constraints. The calculations described here can be used to show that a variant could could only be causal if the prevalence of the disease is higher than published estimates, or its penetrance is below a specified value `r citep("10.1016/j.neuron.2016.09.054")`.

@@

All data required to reproduce the analysis, figures and manuscript (compiled in R) are available at <https://github.com/ImperialCardioGenetics/frequencyFilter>. Curated variant interpretations are deposited in ClinVar under the submission name "HCM\_ExAC\_frequency\_review\_2016". ExAC annotations are available at <https://exac.broadinstitute.org>. Our AF calculator app is located at <http://cardiodb.org/alleleFrequencyApp>, with source code available at <http://github.com/jamesware/alleleFrequencyApp>.

@@

The clinical utility of stringent allele frequency thresholds. (a) The number of predicted protein-altering variants (definition in Methods) per exome as a function of the allele frequency filter applied. A one-tailed 95% confidence interval is used, meaning that variants were removed from consideration if their AC would fall within the top 5% of the Poisson probability distribution for the user's maximum credible AF (x axis). (b) The odds ratio for HCM disease-association against allele frequency. The prevalence of variants in sarcomeric HCM-associated genes (_MYH7_, _MYBPC3_, _TNNT2_, _TNNI3_, _MYL2_, _MYL3_, _TPM1_ and _ACTC1_, analysed collectively) in 322 HCM cases and 852 healthy controls were compared for a range of allele frequency bins, and an odds ratio computed (see **Methods**). Data for each bin is plotted at the upper allele frequency cutoff. Error bars represent 95% confidence intervals. The probability that a variant is pathogenic is much greater at very low allele frequencies.
